---
title: "Factor splines"
author: "Samuel Orso"
date: "`r Sys.Date()`"
output: 
 prettydoc::html_pretty:
  theme: architect
  highlight: github
  toc: true
  df_print: kable
  fig_width: 7
  fig_height: 5
---

## General formualtion of the problem
Let the $d$-dimensional random variable $(X_i)_{i=1}^d$ be expressed as a non-linear
factor model through its cumulative distribution function:$$ F_i(X_i) = h(U,V_i),\quad i =1,\dots,d, $$
where $U\sim\mathcal{U}(0,1)$ and $V_i\sim\mathcal{U}(0,1)$ are independent The unknown function
$h(\cdot)$ is believed to be well approximated by the tensor product spline 
$$ \hat{h}(U,V_i) \equiv s(U,V_i) = \mathbf{B}(U)^T\mathbf{C}\mathbf{B}(V_i), $$ where $\mathbf{B}(x)$ is 
the spline basis for $x$ and $\mathbf{C}$ is a matrix of coefficients. Since
$U$ and $V_i$ are not observed, we propose to estimate the unknown function via
the simulated method of moments: $$ \widehat{\mathbf{C}} = \arg\min_{\mathbf{C}}\lVert\widehat{\mathbf{m}}-\widetilde{\mathbf{m}}(\mathbf{C})\rVert^2_{\Omega}, $$
where $\mathbf{m}$ is a $K$-vector of moments estimated on $\{F_i(X_i)\}_{i=1}^d$ and $\Omega$
is a suitable positive semi-definite matrix of weights. At the end of the procedure,
we obtain an estimate of $s(\cdot)$, $$\hat{s}(U,V_i) = \mathbf{B}(U)^T\widehat{\mathbf{C}}\mathbf{B}(V_i)$$
Note that $\widehat{\mathbf{m}}$ is estimated on the observations where $\widetilde{\mathbf{m}}$
is estimated on pseudo-observations generated from $\hat{s}$. In words,
we try to obtain the best fit to the approximation $s(\cdot)$ of the unknown function $h(\cdot)$.

## Problem at hand
In this example we assume that the true function $h(\cdot)$ is parametric, it is the clayton generator:
$$h(U,V_i) = \left(1+\frac{-\log(V_i)}{F_{\alpha}^{-1}(U)}\right)^{-\alpha}, \alpha>0,$$
where $F_{\alpha}(x)$ is the cdf of a gamma distribution with parameter $(\alpha,1)$.

## Loading packages and functions

```{r,echo=FALSE}
rm(list=ls())
setwd("~/Github/samorso/cosplines/")
load(file="opt_bspline.rda")
load(file="opt_bspline1.rda")
load(file="opt_ispline.rda")
load(file="opt_ispline1.rda")
load(file="opt_ispline2.rda")
load(file="opt_ispline3.rda")
```


```{r,message=FALSE,cache=FALSE}
require(splines2)
require(cosplines)
require(nnls)
```

## Setup
```{r}
n <- 1e3 # sample size for one dimension
d <- 3 # number of dimensions
alpha <- 2.5
set.seed(123L)
z <- runif(n) # latent factor
eps <- matrix(runif(n*d),nc=d) # latent error
x <- clayton(z,eps,alpha) # observations
kn <- c(.03,.2,.5,.8,.97) # knots
df <- 3 # degrees of the B-splines
q <- seq.int(.02,.98,.01) # theoritical quantiles for empirical moments
m_hat <- average_moments(x,q) # empirical moments (first is Spearmann, other are quantile dependences)
B <- 5 # number of bootstrap replicates
```

## B-spline basis
We first try with the B-spline basis.

```{r}
# Best approximation achievable given knots and degrees
xx <- yy <- seq.int(.01,.99,.001)
exp_grid <- expand.grid(xx,yy) 
zz_cl <- matrix(clayton(exp_grid[,1],as.matrix(exp_grid),alpha)[,2],nc=length(xx))
A <- bSpline(xx,knots=kn,degree=df,intercept=F,Boundary.knots=c(0,1))
D <- qr.solve(crossprod(A)) %*% t(A)
coefs_hat <- apply(zz_cl,MARGIN=2,FUN=function(x,M)M%*%x,M=D)
coefs_hat_best <- apply(t(coefs_hat),MARGIN=2,FUN=function(x,M)M%*%x,M=D)

# Estimation
# Use the independent copula as starting points
zz <- outer(xx,yy)
coefs_hat <- apply(zz,MARGIN=2,FUN=function(x,M)M%*%x,M=D)
coefs_hat2 <- apply(t(coefs_hat),MARGIN=2,FUN=function(x,M)M%*%x,M=D)
sv <- c(coefs_hat2) # starting values
```

```{r,eval=FALSE}
opt <- optim(par=c(sv),fn=of_smm,method="Nelder-Mead",M=A,n=n,d=d,q=q,
             m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
opt1 <- optim(par=c(coefs_hat_best),fn=of_smm,method="Nelder-Mead",M=A,n=n,d=d,q=q,
              m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
```

```{r}
C_hat <- matrix(opt$par,nc=ncol(A))
C_hat1 <- matrix(opt1$par,nc=ncol(A))

# Objective function at starting values and optimums
of_smm(sv,A,n,d,q,m_hat,B)
of_smm(c(coefs_hat_best),A,n,d,q,m_hat,B)
of_smm(c(C_hat),A,n,d,q,m_hat,B)
of_smm(c(C_hat1),A,n,d,q,m_hat,B)
```

```{r}
# Visualization
xx <- yy <- seq.int(.01,.99,.03)
exp_grid <- expand.grid(xx,yy) 
zz_cl <- matrix(clayton(exp_grid[,1],as.matrix(exp_grid),alpha)[,2],nc=length(xx))
P <- predict(A,xx)
zz_best <- tcrossprod(P%*%coefs_hat_best,P)
zz_hat <- tcrossprod(P%*%C_hat,P)
zz_hat1 <- tcrossprod(P%*%C_hat1,P)
par(mfrow=c(1,2))
persp(xx,yy,zz_cl,theta=-60,phi=30,expand=.6,ticktype='detailed',main="True function")
persp(xx,yy,t(zz_best),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Best approximation")
persp(xx,yy,t(zz_hat),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (independent copula)")
persp(xx,yy,t(zz_hat1),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (best)")
```

```{r}
# Max relative error
norm(t(zz_best)-zz_cl,"I")/norm(t(zz_best),"I")
norm(t(zz_hat)-zz_cl,"I")/norm(t(zz_hat),"I")
norm(t(zz_hat1)-zz_cl,"I")/norm(t(zz_hat1),"I")
```

## I-spline basis

```{r}
# Best approximation achievable given knots and degrees
xx <- yy <- seq.int(.01,.99,.001)
exp_grid <- expand.grid(xx,yy) 
zz <- matrix(clayton(exp_grid[,1],as.matrix(exp_grid),alpha)[,2],nc=length(xx))
A <- iSpline(xx,knots=kn,degree=df,intercept=F,Boundary.knots=c(0,1))
D <- qr.solve(crossprod(A)) %*% t(A)
coefs_hat <- apply(zz,MARGIN=2,FUN=function(x,M)M%*%x,M=D)
coefs_hat_best <- apply(t(coefs_hat),MARGIN=2,FUN=function(x,M)M%*%x,M=D)
coefs_hat_best_p <- apply(t(coefs_hat),MARGIN=2,FUN=function(x,M)nnls(M,x)$x,M=A) # non-negative ls

# Estimation
# Use the independent copula as starting points
zz <- outer(xx,yy)
coefs_hat <- apply(zz,MARGIN=2,FUN=function(x,M)M%*%x,M=D)
coefs_hat2 <- apply(t(coefs_hat),MARGIN=2,FUN=function(x,M)M%*%x,M=D)
sv <- c(coefs_hat2) # starting values
```

```{r,eval=FALSE}
# Optimization
opt2 <- optim(par=c(sv),fn=of_smm,method="Nelder-Mead",M=A,n=n,d=d,q=q,
              m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
# Optimization using best estimates as starting values
opt3 <- optim(par=c(coefs_hat_best),fn=of_smm,method="Nelder-Mead",M=A,n=n,d=d,q=q,
              m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
# Optimization using best estimates as starting values under non-negative constrains
coefs_hat_best_p[coefs_hat_best_p<=0] <- 1e-3
opt4 <- optim(par=log(c(coefs_hat_best_p)),fn=of_smm2,method="Nelder-Mead",M=A,n=n,d=d,q=q,
              m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
# Optimization using best estimates as starting values under non-negative constrains
coefs_hat_best_p[coefs_hat_best_p>=1] <- 1-1e-3
opt5 <- optim(par=boot::logit(c(coefs_hat_best_p)),fn=of_smm3,method="Nelder-Mead",M=A,n=n,d=d,q=q,
              m_hat=m_hat,B=B,control=list(trace=1,maxit=200*length(sv)))
```

```{r}
C_hat <- matrix(opt2$par,nc=ncol(A))
C_hat1 <- matrix(opt3$par,nc=ncol(A))
C_hat2 <- matrix(exp(opt4$par),nc=ncol(A))
C_hat3 <- matrix(boot::inv.logit(opt5$par),nc=ncol(A))

# Objective function at starting values and optimums
of_smm(sv,A,n,d,q,m_hat,B)
of_smm(c(coefs_hat_best),A,n,d,q,m_hat,B)
of_smm(c(C_hat),A,n,d,q,m_hat,B)
of_smm(c(C_hat1),A,n,d,q,m_hat,B)
of_smm(c(C_hat2),A,n,d,q,m_hat,B)
of_smm(c(C_hat3),A,n,d,q,m_hat,B)
```

```{r,fig.align='center',fig.height=12,fig.width=6,cache=T}
# Visulaization
xx <- yy <- seq.int(.01,.99,.03)
exp_grid <- expand.grid(xx,yy) 
zz_cl <- matrix(clayton(exp_grid[,1],as.matrix(exp_grid),alpha)[,2],nc=length(xx))
P <- predict(A,xx)
zz_best <- tcrossprod(P%*%coefs_hat_best_p,P)
zz_hat <- tcrossprod(P%*%C_hat,P)
zz_hat1 <- tcrossprod(P%*%C_hat1,P)
zz_hat2 <- tcrossprod(P%*%C_hat2,P)
zz_hat3 <- tcrossprod(P%*%C_hat3,P)
par(mfrow=c(3,2))
persp(xx,yy,zz_cl,theta=-60,phi=30,expand=.6,ticktype='detailed',main="True function")
persp(xx,yy,t(zz_best),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Best approximation")
persp(xx,yy,t(zz_hat),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (independent copula)")
persp(xx,yy,t(zz_hat1),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (best)")
persp(xx,yy,t(zz_hat2),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (best) positive constr")
persp(xx,yy,t(zz_hat3),theta=-60,phi=30,expand=.6,ticktype='detailed',main="Estimation (best) (0,1) constr")
```

```{r}
# Max relative error
norm(t(zz_best)-zz_cl,"I")/norm(t(zz_best),"I")
norm(t(zz_hat)-zz_cl,"I")/norm(t(zz_hat),"I")
norm(t(zz_hat1)-zz_cl,"I")/norm(t(zz_hat1),"I")
norm(t(zz_hat2)-zz_cl,"I")/norm(t(zz_hat2),"I")
norm(t(zz_hat3)-zz_cl,"I")/norm(t(zz_hat3),"I")
```


## Some remarks:
- Here we propose to use the independent copula to obtain starting values for $\mathbf{C}$ 
since $U$ and $(V_i)_{i=1}^d$ are not observed. Maybe a different strategy can be 
more optimal (closer to the ``ideal'' starting values).  
- All the results depend on the knot sequences and the selected degree for the splines,
those choices are totally arbitrary up to now.  


